Computational methods in landscape ecology

### 2024-10-18 Manuscript.qmd
Updates review
Small ✍️ updates to intro

format:
  pdf:
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{academicons}
        \usepackage{fontawesome5}
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \usepackage{setspace}
        \renewcommand\Affilfont{\footnotesize}
        \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid#1}}}
        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope#1}}}

@@

\noindent \footnotesize M.H.K. Hesselbarth \orcid{0000-0003-1125-9918} \mail{mhk.hesselbarth@gmail.com}

@@

\noindent All authors contributed particular sections to the manuscript coordinated by MHKH.
Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com)

@@

Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems.
The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology.
Nowadays, advanced computational methods are essential to the field due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems.

@@

This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.

>>>>

format:
  pdf:
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{academicons}
        \usepackage{fontawesome5}
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \usepackage{setspace}
        \renewcommand\Affilfont{\footnotesize}
        \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid\hspace{1mm}#1}}}
        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope\hspace{1mm}#1}}}

@@

\noindent \footnotesize M.H.K. Hesselbarth \orcid{0000-0003-1125-9918} \hspace{5mm} \mail{mhk.hesselbarth@gmail.com}

@@

\noindent All authors contributed particular sections to the manuscript and assisted with the revision. Overall coordination, first and revised draft was led by MHKH.
Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com)

@@

Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems.
The link between spatial heterogeneous patterns and ecological processes is the core concept in the research field of landscape ecology.
Nowadays, advanced computational methods are essential to the field due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems.

@@

This review provides an overview of recent developments in computational methods that in our eyes have advanced the research field of landscape ecology.
We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.

### 2024-10-18 Sections/01-introduction.qmd
Small ✍️ updates to intro

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of executing them (Tab. 1).
In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land-use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

The use of computational science to abstract core mechanisms and research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices] and by extension landscape ecology as such.
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the availability of data increases steadily [@Jaric2020IEcologyHarnessing], and because of the complexity of landscape systems [@Newman2019ScalingComplexity], there is often a need to more advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of creating novel insights in these topics (Tab. 1).
In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
We do not aim to provide a systematic literature review or a general introduction to (computational) landscape ecology, but rather a perspective on recent developments and advances in computational methods. For a set of more comprehensive reviews we refer to @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
